Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2018-03-11 output/breed_paper_revised.Rnw
Fix spelling.

Data were processed and models fit in R, an environment for
statistical computing~\citep{R}. Models were fit using the `rstanarm'
package~\citep{rstanarm}, an interface to Stan, a language for
probabilistic programming~\citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package~\citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to sample to sample variance
and hence is preferable to the 95\% interval~\citep{rstanarm}. To
verify the estimates, the analysis was also conducted using classical
methods, specifically frequentist mixed-effects models fit by maximum
likelihood (see supplemental material for details and model
results). Complete data and code used for this analysis is available
through the Open Science Framework at \url{https://osf.io/6ed5k/}.

@@

The high yielding variety M\=/202 was released in 1985 and became the
most widely planted cultivar for two decades~\citep{mckenzie2014}. At
the time of release, M\=/202 yields were reported at 11.0 Mg
ha\tsups{-1}~\citep{m202} which is almost identical to what we . This
value of 11.0 Mg ha\tsups{-1} is what we also found in this analysis
based on the average experimental yields prior to public release
(Fig. \ref{fig:first_yrs}). Since the release of M\=/202, the rice
yields of newly released varieties cultivars have remained between
10.8 and 11.5 Mg ha\tsups{-1} (Fig. \ref{fig:first_yrs}), with the
exception of M\=/204 and M\=/205 (M\=/204 is no longer grown and M\=/205 has
limited adoaptation range so is only grown in a limited area). The
variety cultivar M\=/206 was released in 2004 and is currently still the
most widely grown rice varietycultivar in California. At the time of
its release, M\=/206 was reported to yield \Sexpr{round(9350*1.12/1000,1)} Mg
ha\tsups{-1} -- similar to what we found based on our analysis
(Fig. \ref{fig:first_yrs}). Based on this premise, it is clear that
improvements in yield potential have not been made over the 33-year
period of this study. In contrast, both~\citet{Omar2014} and
~\citet{mckenzie2014}, report yield potential gains in California rice
systems based on analyses that showed that newly released cultivars
had higher yields than previously released cultivars. Based on their
analyses,~\citet{Omar2014} reported yield gains from the rice breeding
program of 37 to 60 kg ha\tsups{-1} yr\tsups{-1} and
~\citet{mckenzie2014} a yield increase of 25 kg ha\tsups{-1}
yr\tsups{-1}. The simpliest model (Model 2) in this study is similar
to the analyses of~\citet{Omar2014} and ~\citet{mckenzie2014}, and
results in similar findings; yields were estimated to increase by
\Sexpr{only_release_coef} kg ha\tsups{-1} yr\tsups{-1}
(Fig. \ref{fig:wo_erosion}), almost identical to estimates by
~\citet{mckenzie2014}.

The only way to reconcile the two apparently contradictory findings
above is that the yield of potential of cultivars decline over time
such that when newly released cultivars are compared with the older
cultivar (as shown by ~\citet{Omar2014}, ~\citet{mckenzie2014}, and
Fig. \ref{fig:wo_erosion}), yields of the new cultivar are
higher. Here we provide evidence that cultivar yield does decline - on
average at a rate of \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1}
(Fig. \ref{fig:RES_year}). When these declines are taken into account,
there is only a
\Sexpr{round(sum((raw_samples$release_yr_c*scl)>=25)/nrow(raw_samples)*100,1)}\%
probability of a rate of yield increase equal to or greater than 25
kg ha\tsups{-1} yr\tsups{-1} during the study
period. ~\citet{Peng2010} also found that a cultivar's yield potential
declines over time and reported a 15\% decline over a 30 yr period.
In this study, over a 30 year period and a \Sexpr{yrs_coef*-1} kg
ha\tsups{-1} yr\tsups{-1} rate, the yield decline would be
approximately \Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg),2)*100}\%.

This analysis suggests that the CCRRF breeding program has
maintained but not substantially improved the yield potential in the
face of changing biotic and abiotic pressures over the 33 year time
frame of this study. This conclusion holds
true even when evaluating the cultivars which have been breed for
broad adaption and high yield potential (M\=/202, M\=/206, and
M\=/105).

@@

Increasing grain yield is essential to meeting the challenges of
feeding a growing population in a changing world. Here we show that
side-by-side cultivar comparisons do not necessarily indicate gains in
yield potential as they do not account for yield declines over time of
the older cultivars being tested. Here we show, as have others, that
the yield of a cultivar declines over time. Therefore, quantifying
changes in yield potential requires comparing yields of new cultivar
yields to yields of the older cultivars when they were first
released. Unfortunately, our analysis did not allow us to identify the
cause of yield declines in this case. While efforts should be made to
identify the cause of yield decline over time, this study highlights
both the need for plant breeding to simply maintain current yields, as
well as the challenge of increasing genetic yield potential,
especially in systems such as CA where yield is approaching the
physiological limit. The goal of increasing yield may require the
yield performance of cultivars to be assessed earlier in the selection
process to avoid discarding higher yielding entries without full
consideration. We suggest here that historical benchmarks be used in
to evaluate promising rice cultivars in addition to side-by-side
comparisons to maintain perspective of the absolute yield gains being
made.

>>>>

Data were processed and models fit in R, an environment for
statistical computing~\citep{R}. Models were fit using the `rstanarm'
package~\citep{rstanarm}, an interface to Stan, a language for
probabilistic programming~\citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package~\citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to sample-to-sample variance
and hence is preferable to the 95\% interval~\citep{rstanarm}. To
verify the estimates, the analysis was also conducted using classical
methods, specifically frequentist mixed-effects models fit by maximum
likelihood (see supplemental material for details and model
results). Complete data and code used for this analysis is available
through the Open Science Framework at \url{https://osf.io/6ed5k/}.

@@

The high yielding variety M\=/202 was released in 1985 and became the
most widely planted cultivar for two decades~\citep{mckenzie2014}. At
the time of release, M\=/202 yields were reported at 11.0 Mg
ha\tsups{-1}~\citep{m202} which is almost identical to what we . This
value of 11.0 Mg ha\tsups{-1} is what we also found in this analysis
based on the average experimental yields prior to public release
(Fig. \ref{fig:first_yrs}). Since the release of M\=/202, the rice
yields of newly released varieties cultivars have remained between
10.8 and 11.5 Mg ha\tsups{-1} (Fig. \ref{fig:first_yrs}), with the
exception of M\=/204 and M\=/205 (M\=/204 is no longer grown and M\=/205 has
limited adaptation range so is only grown in a limited area). The
variety cultivar M\=/206 was released in 2004 and is currently still the
most widely grown rice cultivar in California. At the time of
its release, M\=/206 was reported to yield \Sexpr{round(9350*1.12/1000,1)} Mg
ha\tsups{-1} -- similar to what we found based on our analysis
(Fig. \ref{fig:first_yrs}). Based on this premise, it is clear that
improvements in yield potential have not been made over the 33-year
period of this study. In contrast, both~\citet{Omar2014} and
~\citet{mckenzie2014}, report yield potential gains in California rice
systems based on analyses that showed that newly released cultivars
had higher yields than previously released cultivars. Based on their
analyses,~\citet{Omar2014} reported yield gains from the rice breeding
program of 37 to 60 kg ha\tsups{-1} yr\tsups{-1} and
~\citet{mckenzie2014} a yield increase of 25 kg ha\tsups{-1}
yr\tsups{-1}. The simplest model (Model 2) in this study is similar
to the analyses of~\citet{Omar2014} and ~\citet{mckenzie2014}, and
results in similar findings; yields were estimated to increase by
\Sexpr{only_release_coef} kg ha\tsups{-1} yr\tsups{-1}
(Fig. \ref{fig:wo_erosion}), almost identical to estimates by
~\citet{mckenzie2014}.

The only way to reconcile the two apparently contradictory findings
above is that the yield of potential of cultivars decline over time
such that when newly released cultivars are compared with the older
cultivar (as shown by ~\citet{Omar2014}, ~\citet{mckenzie2014}, and
Fig. \ref{fig:wo_erosion}), yields of the new cultivar are
higher. Here we provide evidence that cultivar yield does decline - on
average at a rate of \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1}
(Fig. \ref{fig:RES_year}). When these declines are taken into account,
there is only a
\Sexpr{round(sum((raw_samples$release_yr_c*scl)>=25)/nrow(raw_samples)*100,1)}\%
probability of a rate of yield increase equal to or greater than 25
kg ha\tsups{-1} yr\tsups{-1} during the study
period. ~\citet{Peng2010} also found that a cultivar's yield potential
declines over time and reported a 15\% decline over a 30 yr period.
In this study, over a 30-year period and a \Sexpr{yrs_coef*-1} kg
ha\tsups{-1} yr\tsups{-1} rate, the yield decline would be
approximately \Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg),2)*100}\%.

This analysis suggests that the CCRRF breeding program has
maintained but not substantially improved the yield potential in the
face of changing biotic and abiotic pressures over the 33-year time
frame of this study. This conclusion holds
true even when evaluating the cultivars which have been breed for
broad adaption and high yield potential (M\=/202, M\=/206, and
M\=/105).

@@

Increasing grain yield is essential to meeting the challenges of
feeding a growing population in a changing world. Here we show that
side-by-side cultivar comparisons do not necessarily indicate gains in
yield potential as they do not account for yield declines over time of
the older cultivars being tested. Here we show, as have others, that
the yield of a cultivar declines over time. Therefore, quantifying
changes in yield potential requires comparing yields of new cultivar
yields to yields of the older cultivars when they were first
released. Unfortunately, our analysis did not allow us to identify the
cause of yield declines in this case. While efforts should be made to
identify the cause of yield decline over time, this study highlights
both the need for plant breeding to simply maintain current yields,
and the challenge of increasing genetic yield potential, especially in
systems such as CA where yield is approaching the physiological
limit. The goal of increasing yield may require the yield performance
of cultivars to be assessed earlier in the selection process to avoid
discarding higher yielding entries without full consideration. We
suggest here that historical benchmarks be used in to evaluate
promising rice cultivars in addition to side-by-side comparisons to
maintain perspective of the absolute yield gains being made.
